---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

# vpfkit

<!-- badges: start -->
[![R-CMD-check](https://github.com/deng-lab/vpfkit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/deng-lab/vpfkit/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**vpfkit** ("ViroProfiler toolkit") turns the output of the
[ViroProfiler](https://github.com/deng-lab/viroprofiler) viral metagenomics
pipeline into something you can analyse in R. It reads the per-tool result
tables, assembles them into a single `TreeSummarizedExperiment` (TSE), and
gives you three ways to work with that object: an interactive Shiny viewer, a
rendered HTML quality report, and plain-text exports.

```{r overview, echo = FALSE, fig.width = 9, fig.height = 4.2, fig.cap = "ViroProfiler writes one table per tool; create_vpftse() assembles them into a single TreeSummarizedExperiment, which the viewer, the report and the exporters all read."}
library(ggplot2)

box <- function(x, y, w, h, label, fill, colour = "#1F2933", size = 3.1) {
  list(
    annotate("rect", xmin = x - w/2, xmax = x + w/2, ymin = y - h/2, ymax = y + h/2,
             fill = fill, colour = colour, linewidth = 0.4),
    annotate("text", x = x, y = y, label = label, size = size, colour = colour,
             lineheight = 0.95)
  )
}
arrow_style <- arrow(length = unit(0.18, "cm"), type = "closed")

ggplot() +
  box(1.1, 3, 1.9, 3.6,
      "ViroProfiler\npipeline\n\nCheckV\ngeNomad\nVirSorter2\nCoverM\nDRAM-v\nPharokka",
      "#E3F2FD") +
  box(4.0, 3, 1.9, 1.5, "create_vpftse()\nbatch_create_vpftse()", "#B2DFDB") +
  box(4.0, 1.0, 1.9, 0.9, "TreeSummarized-\nExperiment", "#FFE0B2") +
  box(7.2, 4.5, 2.3, 0.9, "run_app()\ninteractive viewer", "#F3E5F5") +
  box(7.2, 3.0, 2.3, 0.9, "generate_report()\nHTML report", "#F3E5F5") +
  box(7.2, 1.5, 2.3, 0.9, "export_abundance()\nexport_annotations()", "#F3E5F5") +
  annotate("segment", x = 2.1, xend = 3.0, y = 3, yend = 3,
           arrow = arrow_style, linewidth = 0.5) +
  annotate("segment", x = 4.0, xend = 4.0, y = 2.2, yend = 1.5,
           arrow = arrow_style, linewidth = 0.5) +
  annotate("segment", x = 5.0, xend = 6.0, y = 1.2, yend = c(4.5, 3.0, 1.5),
           arrow = arrow_style, linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 8.6), ylim = c(0.6, 5.2)) +
  theme_void() +
  theme(plot.margin = margin(4, 4, 4, 4))
```

*vpfkit does not run any bioinformatics tool itself. It starts where
ViroProfiler stops: at the result tables.*

## Installation

```r
install.packages("remotes")
remotes::install_github("deng-lab/vpfkit")
```

vpfkit depends on Bioconductor packages, so install BiocManager first if you
do not have it:

```r
install.packages("BiocManager")
BiocManager::install(c("SummarizedExperiment", "TreeSummarizedExperiment",
                       "mia", "miaViz", "scater"))
```

Report generation additionally needs the [Quarto CLI](https://quarto.org/docs/get-started/).
Everything else works without it.

## Building a TSE from a ViroProfiler run

`create_vpftse()` takes one file per tool and returns a single object holding
the abundance matrices, the per-contig annotations and the sample metadata:

```r
library(vpfkit)

tse <- create_vpftse(
  fin_abcount    = "abundance/abundance_contigs_count.tsv.gz",
  fin_abtpm      = "abundance/abundance_contigs_tpm.tsv.gz",
  fin_abtmm      = "abundance/abundance_contigs_trimmed_mean.tsv.gz",
  fin_abcov      = "abundance/abundance_contigs_covered_fraction.tsv.gz",
  fin_taxa       = "taxonomy/taxonomy_tse.tsv",
  fin_checkv     = "checkv/quality_summary.tsv",
  fin_virsorter2 = "virsorter2/final-viral-score.tsv",
  fin_vibrant    = "vibrant/VIBRANT_genome_quality.tsv",
  fin_dvf        = "dvf/contigs.fa_gt1bp_dvfpred.txt",
  fin_replicyc   = "bacphlip/contigs.fa.bacphlip"
)
```

`batch_create_vpftse()` does the same for a whole ViroProfiler output
directory. `create_vpftse_vir()` narrows the object down to the contigs that
the detectors agree are viral.

The object carries four abundance assays, all produced by CoverM:

| Assay | What it measures |
|---|---|
| `counts` | Reads mapped per contig per sample |
| `tpm` | Transcripts per million; each column sums to 10^6^ |
| `trimmed_mean` | Trimmed mean of per-base coverage depth, in fold coverage |
| `covfrac` | Covered fraction: the share of the contig covered at least once |

`trimmed_mean` is CoverM's trimmed mean of coverage depth. Objects written by
older versions of vpfkit store it under the name `tmm`; both names are
accepted everywhere in this package. It is **not** edgeR's Trimmed Mean of
M-values, and it is not a normalized count.

`covfrac` measures breadth of coverage, not abundance. Its use is to decide
whether a contig is present at all: a contig that collects reads over one
short conserved stretch, with no coverage elsewhere, is usually a false
positive. Requiring 75% breadth is the common convention (Roux et al., 2017,
*PeerJ* 5:e3817); the ViroProfiler paper uses 50% (Ru et al., 2023,
*Gut Microbes*). Apply the filter to raw counts **before** normalizing.

## The interactive viewer

ViroProfiler-viewer is a Shiny app shipped inside the package. It lets you
filter contigs by length, quality, taxonomy, host and replication cycle, and
inspect the result as tables and plots.

```r
vpfkit::run_app()
```

From a terminal, or on a server:

```sh
R -e "options(shiny.port = 8080, shiny.host = '0.0.0.0'); vpfkit::run_app()"
```

Then open <http://localhost:8080>. Upload a `viroprofiler_output.rds` produced
by the pipeline, or built with `create_vpftse()`.

A container image is available for a deployment that should not depend on the
local R installation:

```sh
docker run --rm -p 3838:3838 denglab/vpfkit:latest
```

The image is built from [docker/Dockerfile](docker/Dockerfile) for both
linux/amd64 and linux/arm64.

## Reports and exports

`generate_report()` renders a self-contained HTML summary of a TSE: dataset
composition, CheckV quality, coverage breadth, taxonomy, diversity, abundance
and gene annotations. Sections whose input is missing explain what is missing
instead of failing, so a two-sample object without gene annotations still
produces a complete report.

```r
generate_report(tse, "reports/quality_report.html")

# Zero out abundances below a coverage-breadth threshold first; the
# threshold used is stated in the report.
generate_report(tse, "reports/filtered.html", covfrac_threshold = 0.75)
```

The export functions write flat files for use outside R:

```r
export_abundance(tse, "abundance_counts.csv")
export_abundance(tse, "abundance.xlsx", assay.type = "tpm", format = "xlsx")
export_annotations(tse, "contig_annotations.tsv")
export_vpftse(tse, "viroprofiler_output.rds")
```

Each of them creates the output directory if it does not exist, and returns
the path it wrote.

## Documentation

- [vignettes/vpfkit.Rmd](vignettes/vpfkit.Rmd) — a worked example from result
  tables to report.
- [inst/report_template.qmd](inst/report_template.qmd) — the Quarto template
  `generate_report()` renders. Copy it and edit if you want a different report.
- [ViroProfiler](https://github.com/deng-lab/viroprofiler) — the pipeline that
  produces the input.

## Citation

Ru J, Khan Mirzaei M, Xue J, Peng X, Deng L. ViroProfiler: a containerized
bioinformatics pipeline for viral metagenomic data analysis.
*Gut Microbes* (2023). doi:10.1080/19490976.2023.2192522

## License

GPL (>= 3). See [LICENSE.md](LICENSE.md).
