---
title: "Getting started with vpfkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with vpfkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vpfkit)
library(SummarizedExperiment)
```

vpfkit reads the result tables of a [ViroProfiler](https://github.com/deng-lab/viroprofiler)
run, assembles them into one `TreeSummarizedExperiment` (TSE), and provides an
interactive viewer, an HTML report and flat-file exports on top of it.

## Building the object

With a real ViroProfiler output directory you would call `create_vpftse()` with
one file per tool, or `batch_create_vpftse()` on the directory as a whole:

```{r create, eval = FALSE}
tse <- create_vpftse(
  fin_abcount    = "abundance/abundance_contigs_count.tsv.gz",
  fin_abtpm      = "abundance/abundance_contigs_tpm.tsv.gz",
  fin_abtmm      = "abundance/abundance_contigs_trimmed_mean.tsv.gz",
  fin_abcov      = "abundance/abundance_contigs_covered_fraction.tsv.gz",
  fin_taxa       = "taxonomy/taxonomy_tse.tsv",
  fin_checkv     = "checkv/quality_summary.tsv",
  fin_virsorter2 = "virsorter2/final-viral-score.tsv",
  fin_vibrant    = "vibrant/VIBRANT_genome_quality.tsv",
  fin_dvf        = "dvf/contigs.fa_gt1bp_dvfpred.txt",
  fin_replicyc   = "bacphlip/contigs.fa.bacphlip"
)

# Everything under one ViroProfiler output directory:
tse <- batch_create_vpftse("path/to/viroprofiler_output")
```

So that this vignette runs anywhere, the rest of it uses a small object with
the same structure, built by hand:

```{r toy}
set.seed(1)
contigs <- paste0("NODE_", 1:12, "_length_8000_cov_6.2")
samples <- paste0("S", 1:4)

mk <- function(f) matrix(f(length(contigs) * length(samples)),
                         nrow = length(contigs),
                         dimnames = list(contigs, samples))

tse <- SummarizedExperiment(
  assays = list(
    counts       = mk(function(n) as.numeric(rpois(n, 40))),
    tpm          = mk(function(n) round(runif(n, 0, 5000), 1)),
    trimmed_mean = mk(function(n) round(runif(n, 0, 30), 2)),
    covfrac      = mk(function(n) runif(n))
  ),
  rowData = S4Vectors::DataFrame(
    checkv_quality       = rep(c("High-quality", "Medium-quality", "Low-quality"), 4),
    checkv_contig_length = seq_len(12) * 1500L,
    checkv_completeness  = seq(15, 98, length.out = 12),
    Family               = rep(c("Peduoviridae", "Autographiviridae", NA), 4),
    row.names            = contigs
  ),
  colData = S4Vectors::DataFrame(
    sample_name = samples,
    group       = rep(c("case", "control"), each = 2),
    row.names   = samples
  )
)
tse
```

## The four abundance assays

ViroProfiler runs CoverM several times and vpfkit keeps four of the resulting
tables. They are not interchangeable:

```{r assays}
assayNames(tse)
```

- `counts` — reads mapped per contig per sample.
- `tpm` — transcripts per million; each column sums to 10^6^.
- `trimmed_mean` — the trimmed mean of per-base coverage depth, in fold
  coverage. Objects written by older versions of vpfkit store this under the
  name `tmm`; both names work throughout the package. It is **not** edgeR's
  Trimmed Mean of M-values.
- `covfrac` — the covered fraction, i.e. the share of the contig covered at
  least once.

## Filtering by coverage breadth

`covfrac` measures breadth, not abundance. A contig can accumulate reads over
one short conserved stretch and have no coverage anywhere else; requiring a
minimum breadth before believing an abundance removes most such false
positives. Roux et al. (2017, *PeerJ* 5:e3817) use 75%; the ViroProfiler paper
(Ru et al., 2023, *Gut Microbes*) uses 50%.

`abundance_adjust_by_covfrac()` applies the mask. Apply it to **raw counts**,
before any normalization: masking an already normalized matrix without
renormalizing leaves every sample with a different residual scaling factor.

```{r mask}
counts_masked <- abundance_adjust_by_covfrac(
  assay(tse, "counts"),
  assay(tse, "covfrac"),
  covfrac_threshold = 0.75
)

c(nonzero_before = sum(assay(tse, "counts") > 0),
  nonzero_after  = sum(counts_masked > 0))
```

## Reports

`generate_report()` renders a self-contained HTML summary. Every section
degrades to an explanatory note when its input is missing, so an object
without taxonomy or gene annotations still produces a complete report rather
than failing.

```{r report, eval = FALSE}
generate_report(tse, "quality_report.html")

# Mask before summarizing; the threshold is stated in the report.
generate_report(tse, "quality_report_filtered.html", covfrac_threshold = 0.75)
```

Report generation needs the [Quarto CLI](https://quarto.org/docs/get-started/).
The template is `system.file("report_template.qmd", package = "vpfkit")`; copy
it and edit if you want a different report.

## Exports

```{r exports}
outdir <- file.path(tempdir(), "vpfkit-vignette")

export_abundance(tse, file.path(outdir, "abundance_counts.csv"))
export_annotations(tse, file.path(outdir, "contig_annotations.tsv"))
export_vpftse(tse, file.path(outdir, "viroprofiler_output.rds"))

list.files(outdir)
```

Each exporter creates the output directory when it does not exist and returns
the path it wrote. `export_abundance()` also writes xlsx, naming the worksheet
after the assay:

```{r export-xlsx, eval = requireNamespace("openxlsx", quietly = TRUE)}
f <- file.path(outdir, "abundance.xlsx")
export_abundance(tse, f, assay.type = "tpm", format = "xlsx")
openxlsx::getSheetNames(f)
```

Contig identifiers are written into a leading `Contig` column, so they survive
formats that have no concept of row names:

```{r roundtrip}
back <- read.csv(file.path(outdir, "abundance_counts.csv"), check.names = FALSE)
identical(back$Contig, rownames(tse))
```

```{r cleanup, include = FALSE}
unlink(outdir, recursive = TRUE)
```

## The interactive viewer

```{r app, eval = FALSE}
run_app()
```

`run_app()` starts ViroProfiler-viewer, which filters contigs by length,
quality, taxonomy, host and replication cycle and shows the result as tables
and plots. Upload a `viroprofiler_output.rds`, or one written by
`export_vpftse()`.
